---
title: "SOC & tornadoes"
output: html_document
date: "2024-01-05"
---

Reproducing the analysis of daily U.S. tornado frequency distributions from https://rpubs.com/jelsner/DailyCounts

Get the latest tornado reports from the SPC
```{r}
download.file(url = "http://www.spc.noaa.gov/wcm/data/1950-2022_actual_tornadoes.csv",
              destfile = here::here("data", "Tornadoes.csv"))
Torn.df <- readr::read_csv(file = here::here("data", "Tornadoes.csv"))
```

```{r}
Torn.df |>
  dplyr::filter(yr >= 2013) |>
  dplyr::group_by(yr) |>
  dplyr::summarize(nT = dplyr::n()) |>
  dplyr::summarize(Avg = mean(nT),
                   IQR = IQR(nT))
```

In the ten-year period 2013–2022 there was an average of 1156 tornadoes per year with an interquartile range of 277 tornadoes.

```{r}
YD.df <- Torn.df |>
  dplyr::filter(yr >= 2013) |>
  dplyr::group_by(date, yr) |>
  dplyr::summarize(nT = dplyr::n(), 
                   nVTor = sum(mag >= 4),
                   nNVTor = sum(mag < 4))
Y.df <- Torn.df |>
  dplyr::filter(yr >= 2013) |>
  dplyr::group_by(yr) |>
  dplyr::summarize(nT = dplyr::n())

YmaxD.df <- YD.df |>
  dplyr::group_by(yr) |>
  dplyr::summarize(nTdays = dplyr::n(),
                   maxTor = max(nT))

YmaxD.df <- YmaxD.df |>
  merge(Y.df, by = "yr") |>
  dplyr::mutate(frac = maxTor/nT)

YmaxD.df |>
  dplyr::summarize(minFrac = min(frac),
                   maxFrac = max(frac),
                   minYr = yr[which.min(frac)],
                   maxYr = yr[which.max(frac)])

min(YmaxD.df$frac) * 100
maxFrac = max(YmaxD3.df$frac) * 100
minYr = YmaxD3.df$Year[which.min(YmaxD3.df$frac)]
maxYr = YmaxD3.df$Year[which.max(YmaxD3.df$frac)]
```


The percentage of the annual count occurring on the day of the year with the most tornadoes ranged from a low of 3.6\% in 2018 to a high of 9.5\% in 2021.

```{r}
Torn.df |>
  dplyr::filter(yr >= 2013) |>
  dplyr::summarize(nT = dplyr::n(),
                   nVTor = sum(mag >= 4),
                   perc = nVTor / nT * 100)
```

Moreover, tornadoes are rated on a categorical damage scale ranging from weak EF0 to violent EF4 and EF5. Of the 11,555 U.S. tornadoes in the database over the 10-year period only 0.34\% were rated as violent.

